Computing likelihoods under the SMC for a general class of ARGs.

### 2024-03-14 paper.tex
update

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
\caption{ARG (top) generated under the SMC and its representation as a
series of local trees (bottom). To ensure the one-to-one correspondance
between both representations a node is recorded for each local tree whenever
a node is encoundered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. Blue (unary) nodes represent recombinant nodes.
The blue arrows shows the familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
Note how the presence of the unary node on a local tree is
indicative of either a past (to the left of the local tree)
or future (to the right) coalescence event.
Both ARG representations can be simplified by removing the recombinant nodes.
In that case the information we have on the time to each of these events is
limited to the windows (blue shaded array) delimited by by the age of
the child node $c$ associated with that lineage and the minimum age of
the parents of the edges connecting to $c$.
}
\label{fig:smc-unary}
\end{figure}

@@

@@

\comment{we can probably move this to the next section}
Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
through time by intersecting the internode intervals with the rate change intervals.
Rate changes along the genome can equally be incorporated (see \ref{sec:slicing}).

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge,

\comment{This makes it sound as if tskit is crucial for this observation.}
we can compute a likelihood
under the SMC for any valid tree sequence. The \tskit library provides a whole range
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space,
as well as reduce the number of tracked samples.
Any such subset of the tree sequence is itself a valid tree sequence
that can again be used to compute a likelihood with.

Our approach therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution on the inferred genealogical history of a sample.

Do this independently for any such slice
Although we can readily incorporate changes in $N_e$ and $r$ 
into the likelihood computation, both in time as well as along the genome,

slicing of the ARG can be used to optimize the parameter of interest
for each slice independently.

Figure \ref{fig:3-arg-slices} demonstrates this idea on a toy example.
We simulated an ARG under human-like parameters and with
two population size changes and subsequently estimated the posterior 
distribution using a uniform prior for each slice indepently.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
\caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000, (5000:\infty):10 000]$. Remaining parameters: L=10kb, 20 diploid samples, r=2.5e-8, 148 edges in ARG). A single $N_e$ value was estimated for each slice of the ARG as defined by the true population size change time points. The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior. Number of edges per time slice (from most recent slice to oldest): [68, 57, 23]. Locally unary nodes were omitted for clarity.}
\label{fig:3-arg-slices}
\end{figure}

>>>>

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG (top) generated under the SMC and its representation as a
series of local trees (bottom). To ensure the one-to-one correspondance
between both representations a node is recorded for each local tree whenever
a node is encoundered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. Blue (unary) nodes represent recombinant nodes.
The blue arrows shows the familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
Note how the presence of the unary node on a local tree is
indicative of either a past (to the left of the local tree)
or future (to the right) coalescence event.
Both ARG representations can be simplified by removing the recombinant nodes.
In that case the information we have on the time to each of these events is
limited to the windows (blue shaded array) delimited by by the age of
the child node $c$ associated with that lineage and the minimum age of
the parents of the edges connecting to $c$.
}
\label{fig:smc-unary}
\end{figure}

@@

Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
through time by intersecting the internode intervals with the rate change intervals.
Rate changes along the genome can equally be incorporated (see \ref{sec:slicing}).

@@

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG in both
time as well as along the genome. Our approach therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution on the inferred genealogical history of a sample.
It also opens up the opportunity for more efficient optimization 
strategies during inference given that each slice of the ARG can be
processed independently.

From a data structure perspective, we can compute a likelihood 
under the SMC for any valid tree sequence. 
The \tskit library provides a whole range
of subsetting operations on tree sequences allowing us to slice
the ARG in time and/or space, as well as reduce the number of
tracked samples. Any such subset of the tree sequence is itself
a valid tree sequence that can again be used to compute a
likelihood for. Figure \ref{fig:3-arg-slices} demonstrates this idea 
on a toy example. We simulated an ARG under human-like parameters and with
two population size changes and subsequently estimated the posterior 
distribution using a uniform prior for each slice independently.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
\caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
(5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples, 
$r=2.5e-8$, $148$ edges in ARG). A single $N_e$ value was estimated for each
slice of the ARG as defined by the true population size change time points.
The inferred posterior distribution (right) using the SMC-likelihood was
obtained using a uniform prior. Number of edges per time slice
(from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
omitted for clarity.}
\label{fig:3-arg-slices}
\end{figure}
